---
title: "To Oil or Not to Oil: An Investigation into Agrabathi and Old Wive's Tales"
author: "Dhiya Tathiah and Keeran Moodley"
format:
  pdf:
    include-in-header:
      - file: preamble.tex
    toc: true
    number-sections: true
    colorlinks: true
execute: 
  echo: false
  eval: true
  output: false
---
\newpage
# Introduction 
This assignment explores the influence of different oil treatments on the burn time of incense sticks through the application of a randomized block design (RBD). This study is structured around 3 distinct treatment types (..) randomly applied across 18 experimental units, with the incense brand serving as a blocking factor. By acknowledging the differences that might arise due to manufacturing or material quality, blocking for brands serves as a crucial step in isolating the true effects of the oil treatments. By structuring the experiment in this way, we aim to minimize the impact of confounding variables, allowing us to attribute any variations in burn time more confidently to the treatments themselves rather than to the intrinsic characteristics of the incense sticks.

# Motivation
There is an old wives tale which hails from the ancient Indian subcontinent and has been told for generations; applying oil to an incense stick will cause it to burn faster. This quick burn time ensured that the smoke created a sacred atmosphere, a well scented home and most importantly, carried your family’s prayers to the almighty deities above. While there can be no doubt about the role of incense sticks (or ‘agarbatti’) in the cultural and spiritual settings of an Indian home, doubt remained about the effectiveness of dipping these sticks into the rich, often fragranced, oils. This study seeks to unravel the truth behind this age-old belief, offering modern families the wisdom to discern whether investing in extra oils for incense sticks truly enhances both their connection to the divine and the speed in which the fragrance emanates throughout their homes. By providing this knowledge, we aim to empower families to make informed decisions and potentially save them from unnecessary expenses if these treatments are found not to significantly extend the burn time of incense sticks.

# Objectives
Our objective is to determine which treatment produces the fastest burn rate by testing three options: a control, sandalwood oil, and coconut oil, and assessing their effects on burn time. We will examine whether these commonly used oils differ from each other (comparison of the sandalwood and coconut oil)  and whether there is a difference with the oils from the control (comparing each oil individually to the control). By blocking for the different brands of incense sticks, we can more confidently deduce the differences between treatments, as the blocks contain homogeneous units. Once the incense sticks are lit and the smoke clears, the last burning stick will reveal whether oils truly influence burn time. By repeating this experiment three times, we ensure that our findings are as robust as the scent that lingers in the air.

# Design and Procedure
This experiment will employ a randomised block design with a single factor - application of oil - of three levels, viz., control (no oil), coconut oil, and castor oil. The experiment will block for heterogeneity of experimental units arising from the use of different brands of Agrabathi viz., Hem, Malarani and Tulasi. The factor levels have been selected as they are oils commonly used in Indian households across the world and are the de facto choices during day to day use. The brands of Agrabathi from which the experimental units are drawn from represent easily found and widely exported brands. 

A pilot study will be conducted to assess the viability of the experimental procedure which is outlined below:

1. Select experimental units from each brand of Agrabathi
2. Randomly assign treatments to the units within each block
3. Apply the relevant treatment in the form of coating the sticks of Agrabathi in the appropriate oil ensuring that there is even and consistent covering
4. Light the Agrabathi sticks at their tip and place them in a sheltered area to burn
5. Record the time taken of the Agrabathi to completely burn

Precise details about the randomisation procedure will be discussed in \todo{Link to the relevant section}.

To reduce variance in the experiment due to external factors several steps will be taken to ensure that the experimental conditions will be kept consistent:

1. The Agrabathi will be burnt in the same area to prevent confounding due to location
2. The Agrabathi will be sheltered from wind and sunlight to prevent confounding due to increased airflow over the flaming tip and increased energy due to the sunlight
3. The blocks will be burnt at 10 minute intervals from each other to reduce confounding due to time. The interval is given to allow for the experimenters to set up and light the Agrabathi. This also allows for the majority of the Agrabathi in each group to burn concurrently to further reduce confounding due to time as well as increase the efficiency of the experiment

 The response variable is the time taken for the Agrabathi to burn given in seconds. The measurement of this was achieved via online stopwatch websites and the data was then manually transcribed.


# Pilot study
The pilot study was run with 18 experimental units and blocks were replicated twice. 

Several difficulties were experienced while conducting the pilot study. Due to the large volume of smoke produced by the Agrabathi as it burnt, the experiment had to be conducted outdoors. This made it difficult to control for environmental factors such as wind, humidity, and sunlight. Additionally it was difficult to determine exactly when the Agrabathi stopped burning and thus there are slight non-systematic errors in the measurements of the burn times due to experimental error. 

The original data is provided in the appendix. A basic exploratory analysis was conducted to analyse the data:

```{r}
#Data wrangling and clean up
pilotBurn <- read.csv("AgrabathiBurnTimePilotData.csv")
pilotBurn$time <- pilotBurn$min*60+pilotBurn$sec
pilotBurn$min<- NULL
pilotBurn$sec <- NULL
pilotBurn$block <- factor(pilotBurn$block,levels=c(1,2,3), labels =c("Hem", "Malarani" , "Tulasi"))
pilotBurn$treat<- factor(pilotBurn$treat, levels=c("A","B","C"), labels=c("No oil", "Coconut oil", "Castor oil"))
```

```{r}
#| output: true
#Descriptive stats
library(pastecs)
descstats<-by(pilotBurn$time, pilotBurn$block, stat.desc)
summaryData<- data.frame()
for (block in descstats){
  block<-unname(block)
  summaryData<-rbind(summaryData, data.frame(median=block[8],mean=block[9],var=block[12],stddev=block[13]))
}
row.names(summaryData)<-c("Hem", "Malarani" , "Tulasi")
library(dplyr)
library(kableExtra)
kbl(summaryData,digits=2, format="latex",booktabs=TRUE)|>
kable_styling(position = "center")

```
\todo{analyse this shit}
# Model and outline of analysis
Dhiya justification of assumption
give later 

# The actual experiment and data
randmisation and generation keeran
clean and set up dfs

# ANOVA
dhiya DO from here down on friday 12pm M202

# Contrasts
bonferrini corrections
keeran

# Interpretation
both

# Conclusion
both

# Appendix 

